The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2017-02-06 manuscript/manuscript.Rnw
clarify that the gene shape is for all genes
clarify superorder and species tree
 - taxonomic orders are now approximately
   at the same level for the major taxonomic
   orders
 - indicate that branches are not to scale

Previous studies of some species (i.e., \human, \mouse, \cow, and
\elephant) have provided some preliminary answers to these questions.
They revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
For example, each superorder has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
independent expansions of the prolactins in \mouse, \carmeli, \galili,
and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
related to the thymus-specific serine protease \ac{PRSS16} in
\elephant, and the expansion of the growth hormone and
chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

@@

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,out.width="90mm",echo=FALSE,message=FALSE,warning=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Spalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=gsub("_"," ",label)),nudge_x=0.05,vjust=0.65,hjust=0,size=4,fontface="italic")+
    geom_text(data=data.frame(x=1,y=1),label="Metatheria",nudge_y=0.05,hjust=0,vjust=0,size=4,fontface="italic")+
    scale_x_continuous(limits=c(0,10))+
    theme(legend.margin=unit(0,"mm"),
          plot.margin=unit(c(0,0,0,0),"mm")
          ) + ggplot2::xlim(0,9)

print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.3,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.3,node=2,width=0.13)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.3,node=3,width=0.13)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.3,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.3,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.3,node=6,width=0.12)
      ## sheep -- this is a male, the sample is female, but there aren't any females in phylopic
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.3,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.3,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.3,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.3,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.3,node=11,width=0.15)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.3,node=12,width=0.13)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.3,node=14,width=0.25)
      ## human
      %>% phylopic("acf1cbec-f6ef-4a82-8ab5-be2f963b93f5",color="#AAAA11",alpha=0.3,node=13,width=0.4)
      )
@ 
\caption{Species tree for organisms studied showing the relationship
  between species. Silhouettes of species were obtained from
  \url{http://phylopic.org}. }
\label{fig:species_tree}
\end{figure}

@@

\caption{Distribution shape of placenta core genes. The vertical axis

@@

  placenta transcriptome if %

>>>>

Previous studies of some species (i.e., \human, \mouse, \cow, and
\elephant) have provided some preliminary answers to these questions.
They revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
For example, each superorder (Euarchontoglires, Laurasiatheria,
Afrotheria, and Xenartha, \cref{fig:species_tree})
has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
independent expansions of the prolactins in \mouse, \carmeli, \galili,
and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
related to the thymus-specific serine protease \ac{PRSS16} in
\elephant, and the expansion of the growth hormone and
chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

@@

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,out.width="90mm",echo=FALSE,message=FALSE,warning=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica:7,((Dasypus_novemcinctus:5,Loxodonta_africana:5)Atlantogenata:1,((Canis_familiaris:4,(Equus_caballus:3,(Sus_scrofa:2,(Ovis_aries:1,Bos_taurus:1):1):1):1)Laurasiatheria:1,(((Spalax_carmeli:1,Spalax_galili:1):1,Mus_Musculus:2)Rodentia:2,((Pan_paniscus:1,Homo_sapiens:1):1,Ateles_fuciceps:2)Primates:2)Euarchontoglires:1)Boreoeutheria:1)Eutheria:1);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=gsub("_"," ",label)),nudge_x=0.05,vjust=0.65,hjust=0,size=4,fontface="italic")+
    geom_text(data=data.frame(x=1,y=1),label="Metatheria",nudge_y=0.05,hjust=0,vjust=0,size=4,fontface="italic")+
    geom_text(data=data.frame(x=3.3,y=c(2,3)),label=c("Xenartha","Afrotheria"),nudge_y=0.05,hjust=0,vjust=c(0,0),size=4,fontface="italic")+
    scale_x_continuous(limits=c(0,10))+
    theme(legend.spacing=unit(0,"mm"),
          plot.margin=unit(c(0,0,0,0),"mm")
          ) + ggplot2::xlim(0,9)
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.3,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.3,node=2,width=0.13)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.3,node=3,width=0.13)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.3,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.3,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.3,node=6,width=0.12)
      ## sheep -- this is a male, the sample is female, but there aren't any females in phylopic
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.3,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.3,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.3,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.3,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.3,node=11,width=0.15)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.3,node=12,width=0.13)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.3,node=14,width=0.25)
      ## human
      %>% phylopic("acf1cbec-f6ef-4a82-8ab5-be2f963b93f5",color="#AAAA11",alpha=0.3,node=13,width=0.4)
      )
@ 
\caption{Species tree for organisms studied showing the relationship
  between species. Branch lengths are not to scale. Silhouettes of
  species were obtained from \url{http://phylopic.org}. }
\label{fig:species_tree}
\end{figure}

@@

\caption{Distribution shape of placenta core genes for all possible
  number of genes. The vertical axis

@@

  placenta transcriptome (for all possible number of genes) if %
